Efficient Computation of Extended Sources

### 2021-05-05 repro/papers/sseprecond/paper.tex
interim

Given a function $\phi$ on $z=z_s$, define $(p^{\pm}_1,\bv^{\pm}_1)$
as the solution in the two half-spaces $z <z_s, z>z_s$ of
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
  \lim_{z \rightarrow z_s} P_sp^{\pm}_1& =& \phi.
\end{eqnarray}
By construction, $p^{\pm}_1$ is continuous across $z=z_s$. The
discontinuity $[v^{\pm}_{1,z}]|_{z=z_s}$ is a linear function of
$\phi$. Define the {\em pressure-to-source map} $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s}
\end{equation}
Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solve the
inhomogeneous system
\begin{eqnarray}
\label{eqn:sawe1}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_1}{\partial t} & = &
                                                        (\Lambda^{\pm}_{z_s}\phi)\delta(z-z_s) - \nabla \cdot \bv^{\pm}_1, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \nabla p^{\pm}_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

@@

From the definitions in the last section and the system \ref{eqn:sawe1},
\begin{equation}
  \label{eqn:pident0}
P_r p^{\pm}_0 = \Pi_0{\cal S}^{\pm}(h_s,0)^T = \Pi_0{\cal
  S}^{\pm}\Pi_0^Th_s,
\end{equation}
and
\begin{equation}
  \label{eqn:pident1}
P_r p^{\pm}_0 \approx \Pi_0{\cal S}^{\pm}_{z_s,z_r} (0,f_s)^T = \Pi_0{\cal
  S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s,
\end{equation}
Similarly, from the relations above and \ref{eqn:sawe2},
\begin{equation}
  \label{eqn:vident0}
  P_r  v^{\pm}_{0,z}  \approx \Pi_1{\cal S}^{\pm}_{z_s,z_r}(h_s,0)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_0^Th_s 
\end{equation}
and
\begin{equation}
  \label{eqn:vident1}
  P_r  v^{\pm}_{0,z}  = \Pi_1{\cal S}^{\pm}_{z_s,z_r}(0,f_s)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s 
\end{equation}
Together with \ref{eqn:lamfh}, these equations imply that
\begin{equation}
  \label{eqn:lamnull}
{\cal  S}_{z_s,z_r}^{\pm}P_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}P_1^T.
\end{equation}
Consequently, ${\cal S}_{z_s,z_r}$ has an ``approximate null space'',
in the sense that its composition with the block operator
$(\Lambda^{\pm}_{z_s},-I)^T$ is asymptotically negligible.

@@

Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
\[
  \Lambda^+_{z_s} P_s p^+ = 2 \lim_{z \rightarrow z_s^+}v^+_z 
\]
and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
\[
  S^+_{z_s,z_r}h_s = P_r p^+
\]
Together with the last two equations, the relation
\ref{eqn:approxinv0} implies that
\begin{equation}
  \label{eqn:approxinv1}
  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) = P_rp^+.
\end{equation}
As shown in the last section, the solution of the system
\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
(ii) $\psi = [p^-_2]|_{z=z_r} = -2 lim_{z \rightarrow z_r^-} p^-_2
\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
\begin{equation}
  \label{eqn:approxinv2}
  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
\end{equation}
Taken together, relations \ref{eqn:approxinv1} and
\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  V^-_{z_r,z_s} S^+_{z_s,z_r} &=& \frac{1}{4}I, \nonumber\\
  S^-_{z_r,z_s} V^+_{z_s,z_r} &=& \frac{1}{4}I.
\end{eqnarray}.
The second equation follows by an exactly analogous argument.

>>>>

Given a function $\phi$ on $z=z_s$, define $(p^{\pm}_1,\bv^{\pm}_1)$
as the solution in the two half-spaces $z <z_s, z>z_s$ of
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
  \lim_{z \rightarrow z_s} P_sp^{\pm}_1& =& \phi.
\end{eqnarray}
By construction, $p^{\pm}_1$ is continuous across $z=z_s$. The
discontinuity $[v^{\pm}_{1,z}]|_{z=z_s}$ is a linear function of
$\phi$. Define the {\em pressure-to-source map} $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s}
\end{equation}
Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solves the
inhomogeneous system
\begin{eqnarray}
\label{eqn:sawe1}
  \frac{1}{\kappa}\frac{\partial p^{\pm}_1}{\partial t} & = &
                                                        (\Lambda^{\pm}_{z_s}\phi)\delta(z-z_s) - \nabla \cdot \bv^{\pm}_1, \nonumber \\
  \rho\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \nabla p^{\pm}_1,\nonumber \\
  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

@@

From the definitions in the last section and the system \ref{eqn:sawe1},
\begin{equation}
  \label{eqn:pident0}
P_r p^{\pm}_0 = \Pi_0{\cal S}^{\pm}(h_s,0)^T = \Pi_0{\cal
  S}^{\pm}\Pi_0^Th_s,
\end{equation}
and
\begin{equation}
  \label{eqn:pident1}
P_r p^{\pm}_0 \approx \Pi_0{\cal S}^{\pm}_{z_s,z_r} (0,f_s)^T = \Pi_0{\cal
  S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s,
\end{equation}
Similarly, from the relations above and \ref{eqn:sawe2},
\begin{equation}
  \label{eqn:vident0}
  P_r  v^{\pm}_{0,z}  \approx \Pi_1{\cal S}^{\pm}_{z_s,z_r}(h_s,0)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_0^Th_s 
\end{equation}
and
\begin{equation}
  \label{eqn:vident1}
  P_r  v^{\pm}_{0,z}  = \Pi_1{\cal S}^{\pm}_{z_s,z_r}(0,f_s)^T  =
  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s 
\end{equation}
Together with \ref{eqn:lamfh}, these equations imply that
\begin{equation}
  \label{eqn:lamnull}
{\cal  S}_{z_s,z_r}^{\pm}\Pi_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}\Pi_1^T.
\end{equation}
Consequently, ${\cal S}_{z_s,z_r}$ has an ``approximate null space'',
in the sense that its composition with the block operator
$(\Lambda^{\pm}_{z_s},-I)^T$ is asymptotically negligible.

@@

Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
\[
  \Lambda^+_{z_s} P_s p^+ = 2 \lim_{z \rightarrow z_s^+}v^+_z 
\]
and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
\[
  S^+_{z_s,z_r}h_s = P_r p^+
\]
Together with the last two equations, the relation
\ref{eqn:approxinv0} implies that
\begin{equation}
  \label{eqn:approxinv1}
  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) = P_rp^+.
\end{equation}
As shown in the last section, the solution of the system
\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
(ii) $\psi = [p^-_2]|_{z=z_r} = -2 lim_{z \rightarrow z_r^-} p^-_2
\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
\begin{equation}
  \label{eqn:approxinv2}
  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
\end{equation}
Taken together, relations \ref{eqn:approxinv1} and
\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & \frac{1}{4}I.\\
\end{eqnarray}.
The second equation follows by an exactly analogous argument, and the
last two by time reversal and interchange of the roles of $z_s$ and
$z_r$.

Having identified approximate two-sided inverses for $S^{\pm},
V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
provides the foundation for practical computations. Applying $\Pi_0$
to both sides of \ref{eqn:lamnull} gives
\[
 \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
 \Lambda^+_{z_s} = S^+{z_s,z_r} \Lambda^+_{z_s}.
\]
From the first approximate identity in \ref{eqn:approxinv},
\begin{equation}
  \label{eqn:lamident}
  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \frac{1}{4}\Lambda^+_{z_s}.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a source for velocity, propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$. I will give a detailed description
of this computation in the section on numerical results.

The importance of this result lies in the failure of the obvious
method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
employ the pressure trace as a source in the velocity equation ($f_s$,
in the notation used above) at $z=z_s$, and read off the velocity
field also at $z=z_s$. This difficulty is related to the existence of
tangentially propagating waves and the lack of continuity of the trace
operator. The method implicit in equation \ref{eqn:lamident} avoids
this difficulty by propagating the fields a positive distance in $z$:
assuming as always that the causal fields are downgoing, this step
eliminates any tangentially propagating fields from consideration.

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) will be important in the
construction of a preconditioner. The relation \ref{eqn:lamident} also
provides an computation for this operator. The idea is to treat the
left-hand side of equation \ref{eqn:lamident} as an approximation to
$\Lambda$, and compute its exact transposes via the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx \frac{1}{4}(\Lambda^+_{z_s})^T.
\end{equation} 
The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
simply be substituted for $\Lambda$ and its transpose in the
preconditioner construction: preconditioners must be (precisely)
self-adjoint, but need only be approximated to perform well.
